Short paper : Automatic predicate testing in formal certification

### 2016-02-21 ProofsAndTests/paper-TAP/content/1-introduction.tex
still a bit more of cutting to do

There are multiple ways for increasing confidence in the software we build. One of them is to formally prove its correctness, using a proof assistant. These proofs assistants enable to write code, logical statements and proof in the same language, and offer the guarantee that every proof will be automatically checked. The appearance of these languages, and the way to write programs in them, varies from one to another. Many of them are functional programming languages, like Coq, Idris and Agda, and others, like the B-Toolkit belong to the imperative paradigm. These different paradigm are internally supported by different logic. Systems like Coq, Idris and Agda are based on various higher order logic (CoC, a variant of ML and LUO respectively) and are realisations of the Curry-Howard correspondence, while the formal B method is based on Hoare logic. These different foundations lead to different philosophies and different ways to implement and verify a software, but all of them greatly increase the confidence of the produced software. However, these guarantee tend to be too often consider as perfect, when they are in fact far from it. Knuth was --certainly ironically-- saying ``Beware of bugs in the above code; I have only proved it correct, not tried it". The reality is precisely that a proof isn't enough\footnote{That's without taking in account that when the proof is only done on the paper (and not machine-checked), the proof can be wrong. The paper never complains about a false proof}. When we prove the correctness of a function, we only gain the guarantee expressed by the proven lemma, and nothing more.

Say we want to implement a formally verified sorting function that works for list of elements of type $T$, where $T$ is ordered by a relation $\leq$.
We can decide to put more or less information inside the type of the sorting function $sort$, in order to make it more or less explicit that this function is precisely a sorting function. We can decide to only use a ``weak" type, like $sort\ :\ \{T:Type\}\ \rightarrow\ List\ T\ \rightarrow\ List\ T$, and to use an external lemma to ensure the correctness of the function. Which property does this function has to respect? First, the output has to be sorted. So first, we need to define this notion of being sorted. For this, we can define an inductive predicate.

@@

The first and second constructor of this predicate says that $[\ ]$ and $[x]$ are sorted according to any order order, and for any $x$. The third one says that a list of two or more elements is sorted if the head $h1$ is lower or equal than the next element $h2$, and if the list deprived from its head is also sorted.

@@

Things more nasty can happen. Imagine that the third constructor of $isSorted$, called $consSorted$ would have been written with a typo, and that the condition $(h1 \leq h2) $ would have been incorrectly written as $(h1 \leq h1)$. Then, the function $sort$ could do whatever it wants, its result would be seen as ``sorted". Said differently, any list is seen as ``sorted" according to this predicate, just because of this single typo. For example, if the bijection condition has been added to the formal specification --and therefore the function can no longer return $[\ ]$ all the time--, it would for example be possible to return the exact input, unchanged. Such a function would also unintentionally be seen as ``correct". One could object that the person doing the proof of correctness should realize that the proof is being done too easily, without having to use the essential property that the output is being built such as any element in the list is always lower or equal than the next element in the list. The reality is that currently, many effort are going in the direction of proof automation, where the machine builds the proof automatically. Many proof assistants are enhanced regularly with new automations that solves automatically some kind of goals. For example, Coq has already a Ring prover~\cite{coq2005}, a solver for Presburger arithmetic, and many others automations, and Idris has been recently equipped with a hierarchy of provers for algebraics structures [ref]. There are even extensions to languages, such as Ltac~\cite{DelahayeLTac} and Mtac~\cite{Ziliani13} that aim to help the automation of tactics.  Thus, the general tendency is precisely to let the machine find the proof on its own, and to only interact with the system when needed. And in the imperative world, the formal B-method also follows the same spirit, with many routine proofs being proven automatically. The problem is that the machine is never going to find a proof ``too easy", and will never report that something seems weird with the specification given by the user. We can't reasonably demand that the user of the proof assistant should read all of the automatically generated proof, looking for proofs that seem ``too easy" in order to find weaknesses in the specifications.

Thus, if we want to trust the proven software, we're now forced to believe that there is adequacy between the formal specification --written in some formal logic-- and the informal requirements. A switch has occurred. We used to have to trust code, but we now have to trust logical statements and predicates. But when the specification is too often as complicated as the code, why should we blindly believe in it, when we've first refused to blindly trust the code? Too few people realize that the proving activity has created this silent assumption of adequacy between the formalised specification and the original real requirements, usually written in english. Of course, the proving activity has undeniably increased the confidence in the produced software, and the aim of this paper isn't to say otherwise. Many successful applications of these formal methods, such as a formally verified C compiler \cite{Leroy09} have shown the value of developing a machined-checked proof of correctness. Instead, the primary aim of this paper is to raise awareness of the adequacy concern, and to see how heterogeneous approaches, that mix both proofs and tests, can help to go a step forward in the certification process, in the context of proof assistants based on type theory.
More precisely, we :
\begin{itemize}
	\item Show some basic approaches to the problem of underspecification (section~\ref{sect:naiveApproaches})
	\item Present a way to test in the proof assistant if the predicate effectively captures the notion that we wanted to define, by automatically generating terms, and we automate these tests as far as possible (section~\ref{sect:testingInside})
	\item Present how we can go a step forward by replacing these tests about the predicate by some proofs (section~\ref{sect:aStepForward})
\end{itemize}

>>>>

There are multiple ways for increasing confidence in the software we build. One of them is to formally prove its correctness, using a proof assistant. These proofs assistants enable to write code, logical statements and proofs in the same language, and offer the guarantee that every proof will be automatically checked. The appearance of these languages, and the way to write programs in them, varies from one to another. Many of them are functional programming languages, like Coq, Idris and Agda, and others, like the B-Toolkit belong to the imperative paradigm. These different paradigm are internally supported by different logic. Systems like Coq, Idris and Agda are based on various higher order logic (CoC, a variant of ML and LUO respectively) and are realisations of the Curry-Howard correspondence, while the formal B method is based on Hoare logic. These different foundations lead to different philosophies and different ways to implement and verify a software, but all of them greatly increase the confidence of the produced software. However, these guarantee tend to be too often consider as perfect, when they are in fact far from it. Knuth was --certainly ironically-- saying ``Beware of bugs in the above code; I have only proved it correct, not tried it". The reality is precisely that a proof isn't enough. When we prove the correctness of a function, we only gain the guarantee expressed by the proven lemma, and nothing more.

Say we want to implement a formally verified sorting function that works for list of elements of type $T$, where $T$ is ordered by a relation $\leq$.
We can decide to only use a ``weak" type, like $sort\ :\ \{T:Type\}\ \rightarrow\ List\ T\ \rightarrow\ List\ T$, and to use an external lemma to ensure the correctness of the function. Which property does this function has to respect? First, the output has to be sorted. So first, we need to define this notion of being sorted. For this, we can define an inductive predicate.

@@

The first and second constructor of this predicate says that $[\ ]$ and $[x]$ are sorted according to any order, and for any $x$. The third one says that a list of two or more elements is sorted if the head $h1$ is lower or equal than the next element $h2$, and if the list deprived from its head is also sorted.

@@

Things more nasty can happen. Imagine that the third constructor of $isSorted$, called $consSorted$ would have been written with a typo, and that the condition $(h1 \leq h2) $ would have been incorrectly written as $(h1 \leq h1)$. Then, the function $sort$ could do whatever it wants, its result would be seen as ``sorted". Said differently, any list would be seen as ``sorted" according to this predicate, just because of this single typo. For example, if the bijection condition has been added to the formal specification --and therefore the function can no longer return $[\ ]$ all the time--, it would for example be possible to return the exact input, unchanged. Such a function would also unintentionally be seen as ``correct". One could object that the person doing the proof of correctness should realize that the proof is being done too easily, without having to use the essential property that the output is being built such as any element in the list is always lower or equal than the next element in the list. The reality is that currently, many effort are going in the direction of proof automation, where the machine builds the proof automatically. Many proof assistants are enhanced regularly with new automations that solves automatically some kind of goals. For example, Coq has already a Ring prover~\cite{coq2005}, a solver for Presburger arithmetic, and many others automations, and Idris has been recently equipped with a hierarchy of provers for algebraics structures [ref]. There are even extensions to languages, such as Ltac~\cite{DelahayeLTac} and Mtac~\cite{Ziliani13} that aim to help the automation of tactics.  Thus, the general tendency is precisely to let the machine find the proof on its own, and to only interact with the system when needed. The problem is that the machine is never going to find a proof ``too easy", and will never report that something seems weird with the specification given by the user. We can't reasonably demand that the user of the proof assistant should read all of the automatically generated proof, looking for proofs that seem ``too easy" in order to find weaknesses in the specifications.

Thus, if we want to trust the proven software, we're now forced to believe that there is adequacy between the formal specification --written in some formal logic-- and the informal requirements. A switch has occurred. We used to have to trust code, but we now have to trust logical statements and predicates. But when the specification is too often as complicated as the code, why should we blindly believe in it, when we've first refused to blindly trust the code? The proving activity has created this silent assumption of adequacy between the formalised specification and the original real requirements, usually written in english. Of course, the proving activity has undeniably increased the confidence in the produced software, and the aim of this paper isn't to say otherwise. Many successful applications of these formal methods, such as a formally verified C compiler \cite{Leroy09} have shown the value of developing a machined-checked proof of correctness. Instead, the primary aim of this paper is to raise awareness of the adequacy concern, and to see how heterogeneous approaches, that mix both proofs and tests, can help to go a step forward in the certification process, in the context of proof assistants based on type theory.
More precisely, we :
\begin{itemize}
	\item Show some basic approaches to the problem of underspecification (section~\ref{sect:naiveApproaches})
	\item Present a way to test in the proof assistant if the predicate effectively captures the notion that we wanted to define, by automatically generating terms, and we automate these tests as far as possible (section~\ref{sect:testingInside})
	\item Present how we can go a step forward by replacing these tests about the predicate by some proofs (section~\ref{sect:aStepForward})
\end{itemize}

### 2016-02-21 ProofsAndTests/paper-TAP/content/2-specCanGoWrong.tex
still a bit more of cutting to do

When confronted to this problem of adequacy between the intuitive notion and the formalised one, a first possibility that usually requires quite a lot of effort is to formalise the notion multiple times, with different predicates, and to prove that they are equivalent.
With our example, that means that we need to find another formalisation $isSorted'$ of being sorted, and to prove the following lemma.
$pred\_equiv : \forall\ (T:Type)\ (l:List\ T), (isSorted\ l) \leftrightarrow (isSorted'\ l)$.
This approach aims at increasing the confidence in our formal definitions by assuming that if we've managed to define multiple times the same notion, then we've surely succeeded to define the notion we wanted. The biggest problem with this approach is to be able to find some alternative formalisations that are sufficiently different from the original one. Obviously, if the new normalisations are too similar to the original one (and in the worst case the new ones are just syntactical variants of the first one), then we won't gain any guarantee. The ideal is to be able to capture the same notion by using very different point of view, but this task can be incredibly difficult, and the more different they are, the harder the equivalence proofs will be.

@@

We could go a step forward by removing the need of doing these specific proofs by hand, because in this case, the predicate isSorted can be decided : there exists an algorithm that decided if a list $l$ is sorted --according to isSorted-- and that produces a proof of $(isSorted\ l)$ if appropriate, or a proof of $(not\ (isSorted\ l))$ otherwise.

\begin{lstlisting}
decideIsSorted : (Tord : Order T) -> (l:List T) -> Dec(isSorted Tord v)
decideIsSorted Tord [] = Yes (NilIsSorted Tord)
decideIsSorted Tord [x] = Yes (SingletonIsSorted Tord x)
decideIsSorted Tord (h1::(h2::t)) with (lowerEqDec Tord h1 h2)
  decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          with (decideIsSorted Tord (h2::t))
    decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          | (Yes h2_tail_sorted) = Yes (ConsSorted h1 h2 t h2_tail_sorted h1_lower_h2)
    decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          | (No h2_tail_not_sorted) = No [...]
  decideIsSorted Tord (h1::(h2::t)) | No h1_not_lower_h2 = No [...]
\end{lstlisting}

When the list has two or more elements, we start by checking if the first two elements are ordered. If so, we continue to check recursively with the list deprived from its head $h1$, and otherwise we already know that the list isn't sorted.

Now, in order to do \emph{tests by proof} on the predicate $isSorted$, we can simply run the above decision procedure.

\begin{lstlisting}
isSorted_test1' : Dec (isSorted natIsOrdered [3,5,7])
isSorted_test1' = decideIsSorted natIsOrdered [3,5,7] 
\end{lstlisting}

And if we ask to evaluate $isSorted\_test1'$, the system will answer $Yes$ and the automatically generated proof of $isSorted\ [3,\ 5,\ 7]$, which is what we want, and we therefore know that the predicate has passed this test.

With this technique, we can run semi-automatically a few tests on the predicate $isSorted$. It is semi-automatic in the sense that we still have to define by hand some terms that we know sorted and unsorted --according to our intention-- but we can let the machine produce the proof that the predicate holds or not on these specific values. This isn't bad --and this is in fact all that is done usually, when it is actually done-- but we would like to have a stronger guarantee, and not only that the predicate will coincide with our intuitive notion on a couple of tested terms. We would like to make sure that our intuitive notion and the formalised one coincide for every term.

>>>>

When confronted to this problem of adequacy between the intuitive notion and the formalised one, a first possibility that usually requires quite a lot of effort is to formalise the notion multiple times, with different predicates, and to prove that they are equivalent.
With our example, that means that we need to find another formalisation $isSorted'$ of being sorted, and to prove the following lemma.
$pred\_equiv : \forall\ (T:Type)\ (l:List\ T), (isSorted\ l) \leftrightarrow (isSorted'\ l)$.
This approach aims at increasing the confidence in our formal definitions by assuming that if we've managed to define multiple times the same notion, then we've surely succeeded to define the notion we wanted. The biggest problem with this approach is to be able to find some alternative formalisations that are sufficiently different from the original one. Obviously, if the new normalisations are too similar to the original one (and in the worst case the new ones are just syntactical variants of the first one), then we won't gain any guarantee. The ideal would be to capture the same notion by using very different point of view, and we will show in section~\ref{sect:testingInside} an original approach for doing so.

@@

We could go a step forward by removing the need of doing these specific proofs by hand, because in this case, the predicate isSorted can be decided : there exists an algorithm that decides if a list $l$ is sorted --according to isSorted-- and that produces a proof of $(isSorted\ l)$ if appropriate, or a proof of $(not\ (isSorted\ l))$ otherwise.

\begin{lstlisting}
decideIsSorted : (Tord : Order T) -> (l:List T) -> Dec(isSorted Tord v)
decideIsSorted Tord [] = Yes (NilIsSorted Tord)
decideIsSorted Tord [x] = Yes (SingletonIsSorted Tord x)
decideIsSorted Tord (h1::(h2::t)) with (lowerEqDec Tord h1 h2)
  decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          with (decideIsSorted Tord (h2::t))
    decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          | (Yes h2_tail_sorted) = Yes (ConsSorted h1 h2 t h2_tail_sorted h1_lower_h2)
    decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          | (No h2_tail_not_sorted) = No [...]
  decideIsSorted Tord (h1::(h2::t)) | No h1_not_lower_h2 = No [...]
\end{lstlisting}
Now, in order to do \emph{tests by proof} on the predicate $isSorted$, we can simply run the decision procedure.

\begin{lstlisting}
isSorted_test1' : Dec (isSorted natIsOrdered [3,5,7])
isSorted_test1' = decideIsSorted natIsOrdered [3,5,7] 
\end{lstlisting}
And if we ask to evaluate $isSorted\_test1'$, the system will answer $Yes$ and the automatically generated proof of $isSorted\ [3,\ 5,\ 7]$, which is what we want, and we therefore know that the predicate has passed this test.

With this technique, we can run semi-automatically a few tests on the predicate $isSorted$. It is semi-automatic in the sense that we still have to define by hand some terms that we know sorted and unsorted --according to our intention-- but we can let the machine produce the proof that the predicate holds or not on these specific values. This isn't bad --and this is in fact all that is done usually, when it is actually done-- but we would like to have a stronger guarantee, and not only that the predicate will coincide with our intuitive notion on a couple of tested terms.

### 2016-02-21 ProofsAndTests/paper-TAP/content/3-testingInside.tex
still a bit more of cutting to do

We want to check that this function and the predicate agree, which means that any list generated by $generateSortedList$ should be a yes instance of the predicate $isSorted$, and reciprocally.

Since the predicate $isSorted$ is decidable with $decideIsSorted$, we can automatically check whether the generated sorted lists of size $n$ are automatically sorted (according to $isSorted$) by running the decision procedure. But since there can be an infinite amount of generated sorted lists of size $n$ when $n > 0$, we will only check that the generator and the predicate coincide on a finite observation\footnote{A finite observation of a stream, also called approximation at rank $m$ of a stream, is a vector of size $m$ that has the same $m$ first elements than the stream} of the resulting stream. The biggest this observation is, the better the guarantee is about $isSorted$.

We show how the observation is made on an example. Let $T$ be a type that only contains three values $A$, $B$ and $C$. Since $T$ is finite, it is therefore obviously recursively enumerable. We define on it the strict order $A < B < C$.
Let's automatically generate the first $m$ sorted lists of $T$, of size $n$, by unfolding $m$ times the result of $generateSortedList$.

>>>>

We want to check that this function and the predicate coincide. Since the predicate $isSorted$ is decidable with $decideIsSorted$, we can automatically check whether the generated sorted lists of size $n$ are automatically sorted (according to $isSorted$) by running the decision procedure. But since there can be an infinite amount of generated sorted lists of size $n$ when $n > 0$, we will only check that the generator and the predicate coincide on a finite observation\footnote{A finite observation of a stream, also called approximation at rank $m$ of a stream, is a vector of size $m$ that has the same $m$ first elements than the stream} (arbitrary big) of the resulting stream. The biggest this observation is, the better the guarantee is about $isSorted$.

We show how the observation is made on an example. Let $T$ be a type that only contains three constant values $A$, $B$ and $C$. Since $T$ is finite, it is therefore obviously recursively enumerable. We define on it the strict order $A < B < C$.
Let's automatically generate the first $m$ sorted lists of $T$, of size $n$, by unfolding $m$ times the result of $generateSortedList$.

### 2016-02-21 ProofsAndTests/paper-TAP/content/4-differentKindOfProperties.tex
still a bit more of cutting to do

In order to go a step forward, we can decide to replace the tests on the predicate by proofs. Instead of testing the predicate on a finite subset of the infinitely many sorted lists, as we did in the previous section, we can prove that any of the automatically generated sorted list is provably sorted according to $isSorted$.

>>>>

In order to go a step forward, we can decide to replace the tests on the predicate by proofs. Instead of testing the predicate on a finite subset of the generated sorted lists, as we did in the previous section, we can prove that any of the automatically generated sorted list is provably sorted according to $isSorted$.

### 2016-02-21 ProofsAndTests/paper-TAP/content/5-conclusion.tex
still a bit more of cutting to do

We've presented some new ideas and also reintroduced some known techniques that aim to gain confidence in a predicate inside a proof assistant. Our new technique is based on the automatic generation of terms that should have the desired property. Thus, our goal was to make sure that a generator of terms and the predicate coincide. This adequacy between the predicate and the generator helps to gain confidence in the predicate, because generating examples of something and giving a definition of the same thing are tasks which are sufficiently different that it is very unlikely that they could coincide by luck. The technique presented on section \ref{sect:testingInside} was based on a finite observation of the automatically generated terms processed by a decision procedure. The limitation of this approach is that the predicate has to be decidable. This isn't a big restriction, as a huge portion of the predicates that are used for writing specifications are decidable. We've also shown on an example how these tests can be completely automated. In section \ref{sect:aStepForward} we've presented ideas for going a step forward and replacing these tests on the predicate by some new proofs. These new proofs have the advantage of not requiring the predicate to be decidable. However, they imply that we have to trust these new specifications, which are this time about the predicate itself.

@@

A natural extension of the work presented here is to make it more generic, and to study how proof assistants could help --by providing primitive bricks-- to gain confidence in the formal definitions that we write. A possibility is to provide many robust and generic concepts, like being sorted, so that the user won't have to define his own notion. However, as soon as dependent types are concerned, and when types like lists become indexed by elements like natural numbers, it becomes usually impossible to reuse immediately definitions that were working for the non indexed version.

>>>>

We've presented some new ideas and also reintroduced some known techniques that aim to gain confidence in a predicate inside a proof assistant. Our new technique is based on the automatic generation of terms that should have the desired property. Thus, our goal was to make sure that the predicate and a generator of terms coincide. This adequacy between the predicate and the generator helps to gain confidence in the predicate, because generating examples of something and giving a definition of the same thing are tasks which are sufficiently different that it is very unlikely that they could coincide by luck. The technique presented on section \ref{sect:testingInside} was based on a finite observation of the automatically generated terms processed by a decision procedure. The limitation of this approach is that the predicate has to be decidable. This isn't a big restriction, as a huge portion of the predicates that are used for writing specifications are decidable. We've also shown on an example how these tests can be completely automated. In section \ref{sect:aStepForward} we've presented ideas for going a step forward and replacing these tests on the predicate by some new proofs. These new proofs have the advantage of not requiring the predicate to be decidable. However, they imply that we have to trust these new specifications, which are this time about the predicate itself.

@@

A natural extension of the work presented here is to make it more generic, and to study how proof assistants could help --by providing primitive bricks-- to gain confidence in the formal definitions that we write. A possibility is to provide many robust and generic concepts, like being sorted, so that the user won't have to define his own notion.

### 2016-02-21 ProofsAndTests/paper-TAP/paper-TAP.tex
Structure of the paper
paper continued
still a bit more of cutting to do

@@

\begin{abstract}
The use of formal methods and proof assistants helps to increase the confidence in critical software. However, a formal proof is only a guarantee relative to a formal specification, and not necessary about the real requirements. These real requirements are expressed in a natural language, like english, and there is always a jump when going from an informal specification to a formal specification expressed in a logical theory. Thus, proving the correctness of a piece of software always makes the implicit assumption that there is adequacy between the formalised specification --written with logical statements and predicates-- and the real requirements. Unfortunately, a huge part of the complexity lies precisely in the specification itself, and it is far from obvious that the formal specification says exactly and completely what it should say. Why should we trust more these predicates than the code that we've first refused to trust blindly, leading to these proofs? We show in this paper that the proving activity hasn't replaced the testing activity but has only changed the object which requires to be tested. Instead of testing code, we now need to test predicates. We will show how to identify what needs to be tested in this new context. We also present recent ideas about how to conduct these tests inside the proof assistant on a few examples, and how to automate them as far as possible.

@@

>>>>

\usepackage{listings}

\lstset{
  basicstyle=\ttfamily,
  columns=fullflexible,
  keepspaces=true,
  mathescape
}

@@

\begin{abstract}
The use of formal methods and proof assistants helps to increase the confidence in critical software. However, a formal proof is only a guarantee relative to a formal specification, and not necessary about the real requirements. There is always a jump when going from an informal specification to a formal specification expressed in a logical theory. Thus, proving the correctness of a piece of software always makes the implicit assumption that there is adequacy between the formalised specification --written with logical statements and predicates-- and the real requirements --often written in english--. Unfortunately, a huge part of the complexity lies precisely in the specification itself, and it is far from obvious that the formal specification says exactly and completely what it should say. Why should we trust more these predicates than the code that we've first refused to trust blindly, leading to these proofs? We show in this paper that the proving activity hasn't replaced the testing activity but has only changed the object which requires to be tested. Instead of testing code, we now need to test predicates. We present recent ideas about how to conduct these tests inside the proof assistant on a few examples, and how to automate them as far as possible.

@@

\input{./content/2-specCanGoWrong.tex}

\input{./content/3-testingInside.tex}

\input{./content/4-differentKindOfProperties.tex}

\input{./content/5-conclusion.tex}

\bibliographystyle{splncs03}
\bibliography{bib-TAP}
